---
title: "Data exploration of European Tracking Network data inside European Offshore Wind Farms"
author: "Lotte Pohl, lotte.pohl@ugent.be"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
execute:
  echo: false
  warning: false
  error: false
  output: true
  
---

TODO:
- check with moratorium data -> how to present?
- put the exact concession zone for deployments

# Setup

## workspace

R packages

```{r pgk}
#| echo: true

# 0. data access
# install.packages("pak")
# library(pak)
# pak::pak("inbo/etn")
library(etn)
# library(DBI)
# library(RPostgreSQL)

# 1. data wrangling
library(dplyr)
library(purrr)
library(tidyr)
# library(glue)
library(forcats)
# library(magrittr)

# 2. spatiotemporal
library(lubridate)
library(sf)

# 3. visualisations
library(ggplot2)
library(plotly)
library(leaflet)
library(patchwork)
library(mapview)

# 4. analyses
# library(mgcv)

# 5. misc
library(knitr)
library(here)

```

working directory

```{r wd}
# Set working directory to project root
opts_knit$set(root.dir = here::here())
# suppress all messages
knitr::opts_chunk$set(message = FALSE)
```


Loading functions

```{r fun}
#| echo: true
source("R/helpers.R")
source("R/get_acoustic_detections.R")
source("R/utils_etn.R")
```

Setting paths

```{r paths}
#| echo: true
path_spatial <- "data/spatial"
path_data_raw <- "data_raw"
path_data_raw_etn <- "data_raw/etn"
path_data <- "data"
path_plots <- "docs/plots"
```

Loading data

```{r load-data-into-env}
#| echo: true
load_files_to_env(path = path_spatial, file_ext = "gpkg")
load_files_to_env(path = path_data_raw_etn, file_ext = "rds")
load_files_to_env(path = path_data, file_ext = "rds")
# knitr::knit_exit()
```


## spatial OWF data

```{r data-spatial}

# fix owf polygons
## TODO: put into spatial data script
windfarms_polygons <- windfarms_polygons %>% sf::st_make_valid()

windfarms_polygons_union <- 
  windfarms_polygons %>%
  group_by(country) %>%
  summarise(geom = st_union(geom))

windfarms_polygons_prod <- 
  windfarms_polygons %>%
  dplyr::filter(status == 'Production')
# knitr::knit_exit()


mapview(windfarms_polygons_union) + windfarms_polygons_prod
```

# ETN data

```{r etn-postgres}
# connect to the etn postgres db
# pgdrv <- dbDriver(drvName = "PostgreSQL")
# con_ETNdb <-DBI::dbConnect(pgdrv,
#                    dbname="ETN",
#                    host="pgetn.vliz.be", 
#                    port=5432,
#                    user = Sys.getenv("ETN_USER"),
#                    password = Sys.getenv("ETN_PWD"))
# # troubleshoooting - probably an IP whitelisting issue

```


## deployments of acoustic receivers

```{r etn-deployments}

# # get unique deployment ids
# deployment_ids_owf <- 
#   deployments %>%
#     dplyr::filter(inside_OWF == TRUE) %>%
#     dplyr::select(deployment_id, deploy_date_time, recover_date_time, country) %>% 
#     sf::st_drop_geometry() %>%
#     unique()

# deployment_stations_BE_OWF_slice1 <-
# deployment_stations_BE_OWF %>% 
#   dplyr::filter(deploy_duration_total > median_deploy_duration) %>%
#   dplyr::select(station_name)
# 
# deployment_stations_BE_OWF %>% st_drop_geometry() %>% slice_max(n = 5, order_by = deploy_duration_total) %>% dplyr::select(station_name) %>% pull() -> station_names
```

let's see when and were the OWF deployments were
```{r etn-deployments-owf}
plot_deployments_owf_country <-
  ggplot(data = deployments %>% dplyr::filter(inside_OWF == TRUE)) +
  geom_segment(aes(x = deploy_date_time, y = country, xend = recover_date_time)) +
  labs(title = "acoustic receiver deployments inside OWF by country", x = "", y = "country") +
  theme_minimal()
plot_deployments_owf_country
save_plot(plot_deployments_owf_country, folder = path_plots)

plot_deployments_owf_station <-
  ggplot(data = deployments %>% 
           dplyr::filter(inside_OWF == TRUE & !grepl(",", country)) %>%
           mutate(
             # This creates a factor where levels are sorted by Country, then Station
             station_ordered = factor(station_name, 
                                      levels = rev(unique(station_name[order(country, deploy_date_time)])))
           )) +
  geom_segment(aes(x = deploy_date_time, 
                   y = station_ordered,  # <-- Changed from station_name to station_ordered
                   xend = recover_date_time, 
                   color = country)) +
  labs(title = "acoustic receiver deployments inside OWF by station", 
       x = "", 
       y = "receiver station") +
  theme_minimal() +
  theme(axis.text.y = element_blank())
plot_deployments_owf_station
save_plot(plot_deployments_owf_station, folder = path_plots)

plot_deployments_owf_station_hist <-
  ggplot(data = deployments_station_sum %>% dplyr::filter(inside_OWF == TRUE & !grepl(",", country))) +
  geom_histogram(aes(y = deploy_duration_total %>% as.numeric() / 365)) +
  scale_y_reverse(n.breaks = 15) +
  facet_wrap(~country) +
  labs(title = "deployment duration per receiver station", y = "deployment duration in years") +
  theme_minimal()
plot_deployments_owf_station_hist
save_plot(plot_deployments_owf_station_hist, folder = path_plots)
```

To explore what species have been detected in BE OWF, we get all belgian OWF stations and get the acoustic detections for those.

```{r etn-deployments-owf-BE}
# plot needed? doesn't seem to make much sense
## add total deploy time (in yrs)

plot_deployments_owf_station_BE <-
  ggplot() +
  geom_hline(data = deployment_stations_BE_OWF, aes(yintercept = fct_reorder(station_name %>% as.factor(), first_deployment, .desc = T), color = dist_to_coast_km), linewidth = 2.5) +
  geom_segment(data = deployments %>% dplyr::filter(station_name %in% deployment_stations_BE_OWF$station_name), aes(x = deploy_date_time, y = station_name, xend = recover_date_time, yend = station_name), color = "white", linewidth = 1) +
  # geom_text(data = deployment_stations_BE_OWF, aes(y = station_name, x = deployment_stations_BE_OWF$first_deployment %>% min() - days(30), label = paste((deploy_duration_total %>% as.numeric / 365) %>% round(digits = 1), "y")), color = "white") +
  labs(title = "acoustic receiver deployments inside Belgian OWF by station", x = "", y = "receiver station", color = "dist to coast (km)") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 4),
        legend.position = "left")
# plot_deployments_owf_station_BE
save_plot(plot_deployments_owf_station_BE, folder = path_plots, height = 25)

plot_deployments_owf_station_BE_hist <-
  ggplot() +
  geom_col(data = deployment_stations_BE_OWF, aes(y = fct_reorder(station_name %>% as.factor(), first_deployment, .desc = T), x = deploy_duration_total %>% as.numeric()  / 365, fill = Concession)) +
  # geom_segment(data = deployments %>% dplyr::filter(inside_OWF == TRUE, country == "Belgium"), aes(x = deploy_date_time, y = fct_reorder(station_name %>% as.factor(), deploy_date_time, .desc = T), xend = recover_date_time), color = "white", alpha = ) +
  # geom_text(data = deployment_stations_BE_OWF, aes(y = station_name, x = deployment_stations_BE_OWF$first_deployment %>% min() - days(30), label = paste((deploy_duration_total %>% as.numeric / 365) %>% round(digits = 1), "y")), color = "white") +
  scale_x_continuous(expand = c(0,0)) +
  labs(title = "", x = "total deployment in years", y = "") +
  theme_minimal() +
  theme(axis.text.y = element_blank())
# plot_deployments_owf_station_BE_hist
save_plot(plot_deployments_owf_station_BE, folder = path_plots, height = 25)

(plot_deployments_owf_station_BE + plot_deployments_owf_station_BE_hist) #%>% ggplotly()
```


### summary stats

TODO
ideas: cumulative days of deployment, overall area covered, dist to coast, # of deployments

```{r etn-deployments-summary}
# how many unique stations are there?
deployments %>% 
  dplyr::filter(inside_OWF == TRUE, country == "Belgium") %>% 
  dplyr::select(station_name) %>%
  unique() %>%
  nrow() # TODO make part of the summary stats

# which receiver stations have had hydrophones deployed for the longest?
deployment_stations_BE_OWF_10max <-
  deployment_stations_BE_OWF %>% st_drop_geometry() %>% slice_max(n = 10, order_by = deploy_duration_total)

deployment_stations_BE_OWF_10max %>% 
  dplyr::mutate(ddt_yrs = (deploy_duration_total %>% as.numeric() / 365) %>% round(digits = 2)) %>%
  dplyr::select(station_name, deploy_duration_total, ddt_yrs, , dist_to_coast_km)

plot_deployments_owf_station_BE_duration <-
  ggplot(data = deployment_stations_BE_OWF) +
  geom_col(aes(x = fct_reorder(station_name, deploy_duration_total, .desc = TRUE), 
               y = (deploy_duration_total %>% as.numeric()) / 365, fill = dist_to_coast_km)) +
  scale_y_continuous(n.breaks = 10, expand = c(0,0)) +
  labs(title = "Total deployment duration of receiver stations within Belgian OWF", y = "years of deployment", x = "station name", fill = "dist to coast in km") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5))
# plot_deployments_owf_station_BE_duration
save_plot(plot_deployments_owf_station_BE_duration, path_plots)

plot_deployments_owf_station_BE_duration_10max <-
  ggplot(data = deployment_stations_BE_OWF_10max) +
  geom_col(aes(x = fct_reorder(station_name, deploy_duration_total, .desc = TRUE), 
               y = (deploy_duration_total %>% as.numeric()) / 365, fill = dist_to_coast_km)) +
  scale_y_continuous(n.breaks = 10, expand = c(0,0)) +
  labs(title = "Top 10 deployment duration of receiver stations within Belgian OWF", y = "years of deployment", x = "station name", fill = "dist to coast in km") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1, size = 7))
plot_deployments_owf_station_BE_duration_10max
save_plot(plot_deployments_owf_station_BE_duration_10max, path_plots)

```


## animal projects

Let's check for relevant animal projects
```{r etn-animal-proj}
# the following keywords are considered in R/etn_data_queries.R
animal_proj_keywords <- c("elasmo","elasmobranch", "seabass", "mackerel", "OWF", "VLIZ", "shark", "ray")

```

## animals
TODO: put somewhere else (doesn't follow the flow of thought here)

Animals tagged in projects that have one of the following keywords in their name: `r animal_proj_keywords`:

```{r etn-animals-animproj}

animals_anim_proj_species <-
  animals_anim_proj %>%
  dplyr::group_by(scientific_name) %>%
  dplyr::filter(!scientific_name %in% c(NA,"Built-in")) %>%
  dplyr::summarise(n_individual = dplyr::n(),
                   animal_proj_codes = paste0(animal_project_code %>% unique(), collapse= ", ")) %>%
  dplyr::arrange(desc(n_individual))

animals_anim_proj_species %>% print(n = 50)
```


## detections

query acoustic detections at the acoustic receiver deployments inside OWF
```{r etn-detections-station}
# tests
# test <- get_acoustic_detections2(deployment_id = 3336)
# test$scientific_name %>% unique()

# # try to run biiig query from all stations within BE OWF
# detections_OWF_BE <-
#   etn::get_acoustic_detections(station_name = 
#                                  deployment_stations_BE_OWF %>% 
#                                  dplyr::filter(deploy_duration_total > median_deploy_duration) %>%
#                                  dplyr::select(station_name))


```

Checking the acoustic detections at the stations that have had receivers deployed the longest
```{r etn-detections-longest-deploy-stations}

detections_BE_OWF_10longestdeployments_sum_animal %>%
  dplyr::ungroup() %>%
  dplyr::group_by(scientific_name) %>%
  dplyr::summarise(n_det = sum(n_det),
                   n_ind = dplyr::n(),
                   first_det = min(first_det) %>% as.Date(),
                   last_det = max(last_det) %>% as.Date()) %>%
  dplyr::arrange(desc(n_det)) %>%
  print(n = 50)

plot_detections_BE_OWF_10longestdeployments <-
  ggplot(data = detections_BE_OWF_10longestdeployments_sum) +
  geom_point(aes(x = month, y = scientific_name, size = n_det)) +
  scale_size(transform = "log10", range = c(1, 5)) +
  labs(title = "detections in 10 longest active stations inside BE OWF", x = "", y = "species", size = "n detections") +
  theme_minimal() #+
  # guides(size = 'none')
plot_detections_BE_OWF_10longestdeployments
save_plot(plot_detections_BE_OWF_10longestdeployments, path_plots)
```

abacus plots per species

```{r etn-detections-species}

plot_detections_BE_OWF_seabass <-
  ggplot(data = detections_BE_OWF_seabass_sum) +
  geom_point(aes(y = animal_id %>% as.factor(), x = month, color = dist_to_coast_km, size = n_det)) +
  labs(x = "", title = "seabass detections inside Belgian OWF", y = "animal id", color = "dist to coast in km", size = "det per month") +
  theme_minimal()
plot_detections_BE_OWF_seabass
save_plot(plot_detections_BE_OWF_seabass, folder = path_plots, height = 18)
# plot_detections_BE_OWF_seabass %>% ggplotly()

# smoothhound
plot_detections_BE_OWF_smoothhound <-
  ggplot(data = detections_BE_OWF_smoothhound_sum) +
  geom_point(aes(y = animal_id %>% as.factor(), x = month, color = dist_to_coast_km, size = n_det)) +
  labs(x = "", title = "smoothhound detections inside Belgian OWF", y = "animal id", color = "dist to coast in km", size = "det per month") +
  theme_minimal()
plot_detections_BE_OWF_smoothhound
save_plot(plot_detections_BE_OWF_smoothhound, folder = path_plots, height = 18)
# plot_detections_BE_OWF_smoothhound %>% ggplotly()

# thornbackray
plot_detections_BE_OWF_thornbackray <-
  ggplot(data = detections_BE_OWF_thornbackray_sum) +
  geom_point(aes(y = animal_id %>% as.factor(), x = month, color = dist_to_coast_km, size = n_det)) +
  labs(x = "", title = "thornbackray detections inside Belgian OWF", y = "animal id", color = "dist to coast in km", size = "det per month") +
  theme_minimal()
plot_detections_BE_OWF_thornbackray
save_plot(plot_detections_BE_OWF_thornbackray, folder = path_plots, height = 18)
# plot_detections_BE_OWF_thornbackray %>% ggplotly()

```

Some summary statistics

```{r etn-detections-sum-stats}

```



Seabass has the most detections inside Belgian OWF. So let's look a bit more in detail into those detections

### seabass

```{r detections-seabass}

plot_detections_seabass <- 
  ggplot() +
  # animal tagging
  geom_point(data = animals_seabass, aes(x = release_date_time, y = fct_reorder(animal_id %>% as.factor(), paste(animal_project_code, release_date_time,  sep = "_"), .desc = T), color = animal_project_code), shape = 4, size = 0.5) +
  # all etn detections (sum)
  geom_point(data = detections_seabass_sum, aes(x = month, y = animal_id %>% as.factor(), size = n_det)) +
  # # BE OWF detections
  geom_point(data = detections_BE_OWF_seabass_sum, aes(x = month, y = animal_id %>% as.factor(), size = n_det), color = "grey") +
  # animal tagging
  geom_point(data = animals_seabass, aes(x = release_date_time, y = fct_reorder(animal_id %>% as.factor(), paste(animal_project_code, release_date_time,  sep = "_"), .desc = T), color = animal_project_code), shape = 4, size = 0.5) +
  scale_size(transform = "log10", range = c(0.1, 2)) +
  labs(title = "seabass detections inside/outside OWF", y = "animal id", x = "", color = "animal project code", size = "n detections (per month)") +
  theme_minimal() +
  theme(axis.text.y = element_blank())
plot_detections_seabass

save_plot(plot_detections_seabass, height = 20, folder = path_plots)
```


## tests

```{r tests}
# b <- ggplot(mtcars, aes(wt, mpg)) +
#   geom_point()
# 
# df <- data.frame(x1 = 2.62, x2 = 3.57, y1 = 21.0, y2 = 15.0)
# b +
#  geom_curve(aes(x = x1, y = y1, xend = x2, yend = y2, colour = "curve"), data = df) +
#  geom_segment(aes(x = x1, y = y1, xend = x2, yend = y2, colour = "segment"), data = df)
```
