---
title: "Data exploration of European Tracking Network data inside European Offshore Wind Farms"
author: "Lotte Pohl, lotte.pohl@ugent.be"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---


# Setup

## workspace

R packages

```{r pgk}
#| output: false

# 0. data access
# install.packages("pak")
# pak::pak("inbo/etn")
library(etn)
library(DBI)
library(RPostgreSQL)

# 1. data wrangling
library(dplyr)
library(purrr)
library(tidyr)
library(glue)
# library(magrittr)

# 2. spatiotemporal
library(lubridate)
library(sf)

# 3. visualisations
library(ggplot2)
library(plotly)
library(leaflet)
library(mapview)

# 4. analyses
# library(mgcv)

# 5. misc
library(knitr)
library(here)

```

working directory

```{r wd}
# Set working directory to project root
opts_knit$set(root.dir = here::here())
```


Loading functions

```{r fun}
source("R/helpers.R")
source("R/get_acoustic_detections.R")
source("R/utils_etn.R")
```

Setting paths

```{r var}
path_spatial <- "data/spatial"
path_data_raw <- "data_raw"
path_data <- "data"
path_plots <- "docs/plots"
```

## loading spatial data

```{r data-spatial}
#| output: false

load_files_to_env(path = path_spatial, file_ext = "gpkg")

# fix owf polygons
## TODO: put into spatial data script
windfarms_polygons <- windfarms_polygons %>% sf::st_make_valid()

windfarms_polygons_union <- 
  windfarms_polygons %>%
  group_by(country) %>%
  summarise(geom = st_union(geom))

windfarms_polygons_prod <- 
  windfarms_polygons %>%
  dplyr::filter(status == 'Production')
# knitr::knit_exit()
```

# ETN data

```{r etn-R}
# connect to ETN R package
if(packageVersion('etn') >= "2.3.0"){
  # if 'new' version of etn R package: read .Renviron directly
  readRenviron(".Renviron")
  Sys.getenv("ETN_USER")
  Sys.getenv("ETN_PWD")
}else{
  # if 'old' version: create con object
  con <- etn::connect_to_etn(Sys.getenv("ETN_USER"), Sys.getenv("ETN_PWD"))}
```

```{r etn-postgres}
# connect to the etn postgres db
# pgdrv <- dbDriver(drvName = "PostgreSQL")
# con_ETNdb <-DBI::dbConnect(pgdrv,
#                    dbname="ETN",
#                    host="pgetn.vliz.be", 
#                    port=5432,
#                    user = Sys.getenv("ETN_USER"),
#                    password = Sys.getenv("ETN_PWD"))
# # troubleshoooting - probably an IP whitelisting issue

```


## acoustic receiver deployments

TODO: potentially outsource etn data operations into separate script(s)

```{r etn-deployments}
deployments_raw <-  etn::get_acoustic_deployments()
save_data(deployments_raw, folder = path_data_raw)

deployments <-
  deployments_raw %>%
  # remove deployments with missing lat or lon
  dplyr::filter(!is.na(deploy_latitude) | !is.na(deploy_longitude)) %>%
  # transform lat/lon into geom
  dplyr::mutate(lat = deploy_latitude,
                lon = deploy_longitude) %>%
  sf::st_as_sf(., coords = c("lon", "lat"), crs = 4326) %>%
  # make new col with boolean that is TRUE if inside OWF
  dplyr::mutate(inside_OWF = lengths(sf::st_within(geometry, windfarms_polygons_prod)) > 0) %>%
  sf::st_join(windfarms_polygons_prod, join = sf::st_within) %>% 
  # remove duplicated deployments
  dplyr::distinct(deployment_id, .keep_all = T) %>%
  # fill recover_date_time with current date 
  dplyr::mutate(recover_date_time = dplyr::if_else(recover_date_time %>% is.na(), Sys.time(), recover_date_time))

save_data(deployments, folder = path_data)

# make summary per receiver station/deployment location
deployments_station_sum <-
  deployments %>%
    # get the duration of each deployment
    dplyr::mutate(deploy_duration = difftime(recover_date_time, deploy_date_time, units = "days")) %>%
    # sf::st_drop_geometry() %>%
    dplyr::group_by(station_name) %>%
    dplyr::summarise(deploy_latitude = mean(deploy_latitude),
                     deploy_longitude = mean(deploy_longitude),
                     geometry = st_centroid(st_combine(geometry)),
                     deploy_duration_total = sum(deploy_duration),
                     deployment_ids = paste0(deployment_id, collapse = ", "),
                     inside_OWF = paste0(inside_OWF %>% unique(), collapse = ", "),
                     country = paste0(country %>% unique(), collapse = ", "))
save_data(deployments_station_sum, folder = path_data)

# get unique deployment ids
deployment_ids_owf <- 
  deployments %>%
    dplyr::filter(inside_OWF == TRUE) %>%
    dplyr::select(deployment_id, deploy_date_time, recover_date_time, country) %>% 
    sf::st_drop_geometry() %>%
    unique()

# get station names from inside Belgian OWF
deployment_stations_BE_OWF <-
  deployments_station_sum %>%
    sf::st_drop_geometry() %>%
    dplyr::filter(country == "Belgium" & inside_OWF == TRUE) %>%
    dplyr::select(station_name) %>%
    dplyr::pull()
```

let's see when and were the OWF deployments were
```{r etn-deployments-owf}
plot_deployments_owf_country <-
  ggplot(data = deployments %>% dplyr::filter(inside_OWF == TRUE)) +
  geom_segment(aes(x = deploy_date_time, y = country, xend = recover_date_time)) +
  labs(title = "acoustic receiver deployments inside OWF by country", x = "", y = "country") +
  theme_bw()
plot_deployments_owf_country
save_plot(plot_deployments_owf_country, folder = path_plots)

plot_deployments_owf_station <-
  ggplot(data = deployments %>% dplyr::filter(inside_OWF == TRUE)& !grepl(",", country)) +
  geom_segment(aes(x = deploy_date_time, y = station_name, xend = recover_date_time, color = country)) +
  labs(title = "acoustic receiver deployments inside OWF by station", x = "", y = "receiver station") +
  theme_bw()
plot_deployments_owf_station
save_plot(plot_deployments_owf_station, folder = path_plots)

plot_deployments_owf_station_hist <-
  ggplot(data = deployments_station_sum %>% dplyr::filter(inside_OWF == TRUE & !grepl(",", country))) +
  geom_histogram(aes(x = deploy_duration_total)) +
  facet_wrap(~country) +
  labs(title = "deployment duration per receiver station", x = "deployment duration in days") +
  theme_bw()
plot_deployments_owf_station_hist
save_plot(plot_deployments_owf_station_hist, folder = path_plots)
```

To explore what species have been detected in BE OWF, we get all belgian OWF stations and get the acoustic detections for those.

```{r etn-deployments-owf-BE}

plot_deployments_owf_station_BE <-
  ggplot(data = deployments %>% dplyr::filter(inside_OWF == TRUE, country == "Belgium")) +
  geom_segment(aes(x = deploy_date_time, y = station_name, xend = recover_date_time)) +
  labs(title = "acoustic receiver deployments inside Belgian OWF by station", x = "", y = "receiver station") +
  theme_bw()
plot_deployments_owf_station_BE
save_plot(plot_deployments_owf_station_BE, folder = path_plots)

# how many unique stations are there?
deployments %>% 
  dplyr::filter(inside_OWF == TRUE, country == "Belgium") %>% 
  dplyr::select(station_name) %>%
  unique() %>%
  nrow()
```


### summary stats

TODO
ideas: cumulative days of deployment, overall area covered, # of deployments

## detections

query acoustic detections at the acoustic receiver deployments inside OWF
```{r etn-detections}
# tests
test <- get_acoustic_detections2(deployment_id = 3336)
test$scientific_name %>% unique()

# try to run biiig query from all stations within BE OWF
detections_OWF_BE <-
  etn::get_acoustic_detections(station_name = deployment_stations_BE_OWF)
```


## tests

```{r tests}
# b <- ggplot(mtcars, aes(wt, mpg)) +
#   geom_point()
# 
# df <- data.frame(x1 = 2.62, x2 = 3.57, y1 = 21.0, y2 = 15.0)
# b +
#  geom_curve(aes(x = x1, y = y1, xend = x2, yend = y2, colour = "curve"), data = df) +
#  geom_segment(aes(x = x1, y = y1, xend = x2, yend = y2, colour = "segment"), data = df)
```
